---
title: "Statistical Enrichment of ORFs"
author: "Ryan Wong"
date: '2022-06-09'
output: html_document
---

## Load required packages

```{r Packages}
library(dplyr)
library(tidyr)
library(readr)
library(readxl)
library(writexl)
```

## Input data

Input file will be the genesMatrix_PresenceAbsence file from [link] (http://1002genomes.u-strasbg.fr/files/)  
Modifications were made in the file to curate the data and add S288C.
The data for the TtL hit strains (9) and the MtL hit strains (15) were separated into it's own sheet.
The hits have 4 overlapping strains making the total number of hits 20.

```{r Data loading}

setwd('D:/') #change all paths as needed

ORFs <- read_excel('D:/genesMatrix_PresenceAbsence.xlsx') #all strains

Hits <- read_excel('D:/genesMatrix_PresenceAbsence.xlsx', sheet = 2) #all hits

```

## Determine Core ORFs

Summarize all numerical values in each column to get total representation.
Select only the ORFs that are found in all 1012 strains and define them as Core ORFs.

```{r Core determination}

sumORFs <- ORFs %>% summarise(across(where(is.numeric), sum))

Core <- sumORFs %>% select(where(~ any(. > 1011)))

```

## Determine S288c ORFs

Select only the ORFs that are found in S288c and define them as S288c ORFs.

```{r S288c determination}

S288c <- ORFs[919, ] %>% select(where(~ any(. > 0))) #S288C was added in as REF and when sorted alphabetically by STD_name ends up in row 919. Adjust accordingly.

```

## Reference Library Curation

Remove the hit strains from the library of all other strains to be able to compare if there is enrichment of certain ORFs vs. the rest of the strains.


```{r Library curation}

rownames(ORFs) <- ORFs$STD_name

rownames(Hits) <- Hits$STD_name

RefLib <-subset(ORFs, !rownames(ORFs) %in% rownames(Hits))

```

## Remove Core ORFs

Previously determined Core ORFs will remove them from the reference and all hits.

```{r Core subtraction}

RefLib <- RefLib[, !colnames(RefLib) %in% colnames(Core)]
RefLib <- data.frame(RefLib)

Hits <- Hits[, !colnames(Hits) %in% colnames(Core)]
Hits <- data.frame(Hits)

```

## Count ORF Occurance

Sum the occurrence of each non-core ORF and non-occurance.

```{r ORF Occurrence}

sumRef <- data.frame(t(RefLib %>% summarise(across(where(is.numeric), sum))))
colnames(sumRef) <- "Occr"
sumRef$N_Occr <- 992 - sumRef$Occr

sumH <- data.frame(t(Hits %>% summarise(across(where(is.numeric), sum))))
colnames(sumH) <- "Occr"
sumH$N_Occr <- 20 - sumH$Occr

```

##Fisher Test Enrichment

Do fishers test for each row of summary tables and output p.values to a new table.

```{r Determine Fisher P-values for Enrichment}

i = 1
Fish_H <- data.frame(rownames(sumH))
while (i < 2776)
{
  TH <- rbind(sumH[i, ], sumRef[i, ])
  Fish_TH <- fisher.test(TH, alternative = "greater")
  Fish_H[i,2] <- Fish_TH$p.value
  i = i + 1
}
colnames(Fish_H) <- c("ORF", "P-Enr")

```

##Fisher Test Depleted

Do fishers test for each row of summary tables and output p.values to a new table.

```{r Determine Fisher P-values for Depletion}

i = 1
while (i < 2776)
{
  TH <- rbind(sumH[i, ], sumRef[i, ])
  Fish_TH <- fisher.test(TH, alternative = "less")
  Fish_H[i,3] <- Fish_TH$p.value
  i = i + 1
}
colnames(Fish_H) <- c("ORF","P-Enr", "P-Dpl")

```

## Calculate Fractional Occurrence

Take the number of time an ORF occurs in the dataset and divide by total number of strains in dataset.

```{r Fraction of strains with ORF}

sumRef$Frac <- sumRef$Occr/992

sumH$Frac <- sumH$Occr/20

```

# Put Fractional Occurance into Percent

Multiply fractional occurrence by 100.

```{r Percent of strains with ORF}

sumRef$Pcnt <- sumRef$Frac*100

sumH$Pcnt <- sumH$Frac*100

```

## Calculate if ORF is Enriched or Depleted

Logical argument is it greater or less than in Reference library.

```{r Greater or less than Reference}

sumH$Grtr_Ref <- F
sumH$Grtr_Ref[sumH$Pcnt > sumRef$Pcnt] <- T
sumH$Less_Ref <- F
sumH$Less_Ref[sumH$Pcnt < sumRef$Pcnt] <- T

```

## Merge Occurance with Fisher P-value

```{r P-values}

sumH$P_Enr <- Fish_H[,2]
sumH$P_Dpl <- Fish_H[,3]

```

## P-value significance

Set cut off of 0.05 or lower.

```{r Significance}

sumH$Sig_Enr <- F
sumH$Sig_Enr[sumH$P_Enr < 0.05] <- T
sumH$Sig_Dpl <- F
sumH$Sig_Dpl[sumH$P_Dpl < 0.05] <- T

```

## Add ORF names to Summary Tables

```{r Add ORF names}

sumH$ORFs <- rownames(sumH)
sumH <- sumH[, c(11, 1, 2, 3, 4, 5, 6, 7, 8, 9, 10)]

```

## Calculate Fold Enrichment

```{r Fold Enrichment}

sumH$Fold_Enr <- sumH$Frac/sumRef$Frac

```

## Calculate Odds Ratio

```{r Odds Ratio}

sumH$Odds_Ratio <- ((sumH$Occr/sumH$N_Occr) / (sumRef$Occr/sumRef$N_Occr))

```

## Two-fold Enrichment

```{r two-fold enrichment}

sumH$twofold <- F
sumH$twofold[sumH$Fold_Enr > 2] <- T

```

## Generate Tables of Enriched and Depleted ORFS

```{r Enriched and Depleted ORFs}

Hits_Enriched <- sumH %>% filter(Sig_Enr == T)
Hits_Depleted <- sumH %>% filter(Sig_Dpl == T)
write_xlsx(Hits_Enriched, 'D:/Hits_Enriched_ORFs.xlsx')
write_xlsx(Hits_Depleted, 'D:/Hits_Depleted_ORFs.xlsx')

```

## Generate Tables of Enriched and Depleted ORFS removing those in S288c

```{r S288c ORF Curation}

Hits_Enriched_no_S288c <- Hits_Enriched[!rownames(Hits_Enriched) %in% colnames(S288c), ]
Hits_Depleted_no_S288c <- Hits_Depleted[!rownames(Hits_Depleted) %in% colnames(S288c), ]
Hits_Depleted_S288c <- Hits_Depleted[rownames(Hits_Depleted) %in% colnames(S288c), ]
write_xlsx(Hits_Enriched_non_S288c, 'D:/Hits_Enriched_non-S288c_ORFs.xlsx')
write_xlsx(Hits_Depleted_non_S288c, 'D:/Hits_Depleted_non-S288c_ORFs.xlsx')
write_xlsx(Hits_Depleted_S288c, 'D:/Hits_Depleted_S288c_ORFs.xlsx')

```